---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Benjamin Culberson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
getwd()
library(tidyverse)
library(agricolae)
library(ggplot2)
library(lubridate)

NTL_LTER_raw <- 
  read.csv("../Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", 
           stringsAsFactors = TRUE)

NTL_LTER_raw$sampledate <- as.Date(NTL_LTER_raw$sampledate , format = "%m/%d/%y")

#2

mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")

theme_set(mytheme)


```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: There is no statistically significant relationship between mean lake temperature recorded in July and lake depth across all lakes.
Ha: There is a statistically significant relationship between mean lake temperature recorded in July and lake depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
NTL_LTER_processed <- 
  NTL_LTER_raw %>%
  mutate(Month = month(sampledate)) %>%
  filter(Month == "7") %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  drop_na()

#5
NTL_LTER_temperature_depth <-
  ggplot(NTL_LTER_processed, aes(x = depth, y = temperature_C)) +
  geom_point() +
  ylim(0,35)+
  labs(y = "Temperature (Celsius)", x = "Depth (Meters)", title = "Temperature by Depth")+
  geom_smooth(method = lm, color = "blue")
NTL_LTER_temperature_depth
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: It appears that as depth increases, temperature decreases. The distributions of these points do suggest that this trend is not perfectly linear however. As depth increases past roughly 8-9 meters, temperature appears to remain constant.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
linear_regression_7 <- lm(NTL_LTER_processed$temperature_C ~ NTL_LTER_processed$depth)
summary(linear_regression_7)
plot(linear_regression_7)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: When we perform a linear regression of 'temperature_C' on 'depth' we find that 73.87% of the variability in lake temperature is explained by lake depth with 9726 degrees of freedom. We also find a statistically significant effect of lake depth on temperature at beyond the 0.001 level. My model estimates that for every 1 additional meter of depth, the temperature of the lake's water decreases by 1.95 degrees C. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
TPAIC <- lm(data = NTL_LTER_processed, temperature_C ~ year4 + daynum + depth)

step(TPAIC)

#10
MRmodel <- lm(data = NTL_LTER_processed, temperature_C ~ year4 + daynum + depth)
summary(MRmodel)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggests we use all three explanatory variables to predict temperature in our multiple regression (year, day, and depth). This new multiple regression explains 74.1% of the variation in water temperature. This explained variation is a slight improvement on our linear regression that only uses depth as an explanatary variable (73.87%). 


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
# Format ANOVA as aov
NTL_LTER_processed.anova <- aov(data = NTL_LTER_processed, temperature_C ~ lakename)
summary(NTL_LTER_processed.anova)

NTL_LTER_processed.anova2 <- lm(data = NTL_LTER_processed, temperature_C ~ lakename)
summary(NTL_LTER_processed.anova2)
summary(NTL_LTER_processed$lakename)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: After running an ANOVA test as both an lm() and aov() function, it appears that there is a statistically significant difference in mean temperature among the lakes. The aov() function shows that the mean temperature is statistically different among the lakes below the 0.001 level. However the aov() function does not report the breakdown of which lake is different from the mean. The lm() function does show this breakdown (with the intercept=Central Long Lake mean temperature) and all of the deviations each lake has from the intercept mean temperature. This lm() function too, shows that every single lake's mean temperature is significantly statistically different from the intercept lake's mean temperature (at beyond the 0.001 level). 


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
NTL_LTER_processed_temperature_by_depth <-
  ggplot(NTL_LTER_processed, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point() +
  geom_smooth(method = lm, se= FALSE) + 
  ylim(0,40)+
  labs(y = "Temperature (Celsius)", x = "Depth (meters)", color = "Lake", 
       title = "Temperature by Lake and Depth")
print(NTL_LTER_processed_temperature_by_depth)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(NTL_LTER_processed.anova)

#Let's make it easier to judge the groups
NTL_LTER_processed.anova.groups <- 
  HSD.test(NTL_LTER_processed.anova, "lakename", 
           group = TRUE) 
#group = TRUE means group levels that have the same mean
NTL_LTER_processed.anova.groups

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Based on the findings above, it seems that Paul Lake and Ward Lake have the same mean temperature, statisticallty speaking, as Peter Lake. It also appears that there are no lakes with a mean temperature statistically distinct from all the other lakes. No matter which lake you choose, that lake has the same temperature, statistically speaking, as at least one other lake.
 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We could use a two sample t-test to determine whether Peter Lake and Paul Lake have distinct temperatures, although it does assume that the variances of the two lakes' temperature is equivalent. 




